A highly-available move operation for replicated trees

### 2019-06-20 paper/move-op.tex
paper: more editing throughout
Editing throughout the paper.

For many systems and applications, the data model that describes the system state is a tree.
For example:
\begin{itemize}
    \item The \textbf{filesystem} on most operating systems is organised as a tree: folders or directories are branch nodes, and files are leaf nodes.
        In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
        Strictly speaking, filesystems that support hardlinks form a DAG, but we will focus on trees for now and show in Section~\ref{sec:hardlinks} how to add support for links.
    \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, bulleted and numbered lists, figures, sections, and so on.
        These can be nested inside each other: for example, a paragraph may appear inside a bullet point of a bulleted list, which may in turn appear inside a numbered point of an enumeration.
        Rich text documents are typically represented in memory using the Document Object Model (DOM), and written to file in HTML, XML, or JSON format.
    \item \textbf{Vector graphics} and \textbf{presentation software} represents images using graphical objects such as text boxes, rectangles, ellipses, lines, and so on.
        These objects are contained within nodes representing pages of a document, or slides of a presentation.
        Multiple objects may be combined into a \emph{group} so that they can be manipulated as a unit; this corresponds to making these objects children of a common parent node.
        Multiple objects and groups may in turn be combined into higher-level groups, forming a tree.
    \item Personal \textbf{note-taking} and \textbf{task-management tools} such as Org-mode for Emacs~\cite{OrgMode} or OmniOutliner~\cite{OmniOutliner} explicitly present the user with a tree structure that they can inspect and manipulate.
\end{itemize}

@@

In this paper we use an abstract model of a tree that can support any of the applications listed above, including distributed filesystems.
We allow replicas to manipulate this tree in any way: by creating new nodes, deleting nodes, or moving subtrees to a new location within the tree.
While there are many existing systems that support creating and deleting nodes (see Section~\ref{sec:relwork}), the key innovation of our algorithm is the support for moving subtrees.
We explain in Section~\ref{sec:move-is-hard} why this move operation is so challenging.

Our contributions in this work are as follows:
\begin{itemize}
    \item We define a Conflict-free Replicated Data Type for trees that supports a \emph{move} operation, without requiring any coordination between replicas such as locking or consensus.
        As discussed in Section~\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
    \item We formalise an abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of the algorithm.
        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) while keeping the data in a tree structure.
    \item To demonstrate the practical viability of our approach, we refine our abstract algorithm to an efficient implementation within Isabelle/HOL and prove the equivalence of the two.  We thereafter extract a formally verified, executable Scala implementation from the Isabelle/HOL definitions, and perform tests on this implementation.
    \item We perform experiments with replicated filesystem products such as Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
\end{itemize}

@@

The first difficulty arises when a node is moved to a new location on one replica, and concurrently the same node is moved to a different location on a different replica.
This scenario is illustrated in Figure~\ref{fig:move-same-item}, where replica 1 moves node $A$ to be a child of $B$, while concurrently replica 2 moves $A$ to be a child of $C$.
After the replicas communicate these operations to each other, what should the merged state of the tree be?

@@

In an unreplicated tree it is easy to prevent cycles being introduced: if the node being moved is an ancestor of the destination node, then the move operation must be rejected.
However, a new problem arises in a replicated setting: different replicas may perform operations that are individually safe, but whose combination leads to a cycle.

@@

In this section we walk through the code in Figure~\ref{fig:code} step by step, explaining features of the Isabelle/HOL syntax as we encounter them.
For additional background on the language, standard documentation is available~\cite{DBLP:books/sp/NipkowK14}.

@@

We define only one kind of operation: \isa{Move t p m c} (Figure~\ref{fig:code}, lines 1--5).
A move operation is a record consisting of four fields: a timestamp \isa{t} of type \isa{'t}, a parent node \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node \isa{c} of type \isa{'n}.
Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be \emph{substituted} for arbitrary types; our only requirement is that timestamps \isa{'t} are globally unique and have an associated total order.
Concretely, Lamport timestamps~\cite{Lamport:1978jq} would meet these requirements.

The meaning of an operation \isa{Move t p m c} is that at time \isa{t}, the child node \isa{c} is moved to be a child of parent node \isa{p}.
The operation does not specify the old location of \isa{c}; the algorithm simply ``steals'' \isa{c} from wherever it is currently located in the tree, and moves it to \isa{p}.
If \isa{c} does not currently exist in the tree, it is created as a child of \isa{p}.
For this reason we do not need a separate operation type to create tree nodes: simply performing a move with a fresh child node \isa{c} is sufficient to create \isa{c}.
Moreover, we do not define a remove operation: instead, we can designate a particular tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.

@@

When a replica applies a \isa{Move} operation to its tree, it also records a corresponding \isa{LogMove} operation in its log.
The \isa{t}, \isa{p}, \isa{m} and \isa{c} fields are taken directly from the \isa{Move} record, while the \isa{oldp} field is filled in based on the state of the tree before the move.
If \isa{c} did not exist in the tree prior to the move, \isa{oldp} is set to \isa{None}.
Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}, that is: if there exists a \isa{p'} and \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.
This logic is performed by the \isa{get\_parent} function (lines 16--20).

@@

If the log is nonempty (line 48), we take \isa{logop} to be the first element of the log, and \isa{ops} to be the rest.
(The hash character in \isa{logop {\isacharhash} ops} is the \emph{cons} operator that adds one element to the head of a list.)
If the timestamp of \isa{logop} is greater than the timestamp of the new operation (line 49) we first undo \isa{logop} with \isa{undo\_op}, then recursively apply the new operation to the remaining log, and finally reapply \isa{logop} with \isa{redo\_op} (line 50).
Otherwise we perform the operation using \isa{do\_op}, and add the corresponding \isa{LogMove} record as the head of the log (line 51).

This logic ensures that the log is maintained in descending timestamp order, with the greatest timestamp at the head.
\isa{undo\_op} (lines 33--36) inverts the effect of a previous move operation by restoring the prior parent and metadata that were recorded in the \isa{LogMove}'s additional field.
\isa{redo\_op} (lines 38--41) uses \isa{do\_op} to perform an operation again and recomputes the \isa{LogMove} record (which might have changed due to the effect of the new operation with a lower timestamp).

@@

The first invariant we prove is that each tree node has either no parent (if it is the root of a tree) or exactly one parent (if it is a non-root node).
We state this theorem in Isabelle as follows, where \isa{apply\_ops\_unique\_parent} is the name we give to this theorem:
\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ apply{\isacharunderscore}ops{\isacharunderscore}unique{\isacharunderscore}parent{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isacharequal}\ {\isacharparenleft}log{\isacharcomma}\ tree{\isacharparenright}{\isacartoucheclose}\isanewline
\ \ \isakeyword{shows}\ {\isacartoucheopen}unique{\isacharunderscore}parent\ tree{\isacartoucheclose}
\end{isabelle}
\noindent That is, we consider any list of operations \isa{ops} and define \isa{(log, tree)} to be the replica state after \isa{ops} have been applied.
We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 56--57 of Figure~\ref{fig:code}: it states that whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.

Because we make no assumptions about \isa{ops}, this theorem holds for any replica state that can be reached by applying any number of operations.
Proving \isa{apply\_ops\_unique\_parent} is easy because each of the \isa{do\_op}, \isa{undo\_op} and \isa{redo\_op} functions individually maintains the \isa{unique\_parent tree} invariant, and thus \isa{apply\_op} also maintains the invariant.

@@

To prove that \isa{apply\_op} preserves the acyclicity invariant, we need to show that for any tree produced by \isa{apply\_ops}, there exists a sequence of operations such that applying those operations using \isa{do\_op} invocations alone (not using \isa{undo\_op} or \isa{redo\_op}) results in the same tree.
(This is the case when operations are applied in order of increasing timestamp.)
Then we can use the result of the previous paragraph to prove that any tree produced by \isa{apply\_ops} must be acyclic.

>>>>

Many systems use a tree-structured data model.
For example:
\begin{itemize}
    \item The \textbf{filesystem} on most operating systems is organised as a tree: folders or directories are branch nodes, and files are leaf nodes.
        In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
        Strictly speaking, filesystems that support hardlinks form a DAG, but we will focus on trees for now and show in Section~\ref{sec:hardlinks} how to add support for links.
    \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, bulleted and numbered lists, figures, sections, and so on.
        These can be nested inside each other: for example, a paragraph may appear inside a bullet point of a bulleted list, which may in turn appear inside a numbered point of an enumeration.
        Rich text documents are typically represented in memory using the Document Object Model (DOM), and written to file in HTML, XML, or JSON format.
    \item \textbf{Vector graphics} and \textbf{presentation software} represents images using graphical objects such as text boxes, rectangles, ellipses, lines, and so on.
        These objects are contained within nodes representing pages of a document, or slides of a presentation.
        Multiple objects may be combined into a \emph{group} so that they can be manipulated as a unit; this corresponds to making these objects children of a common parent node.
        Multiple objects and groups may in turn be combined into higher-level groups, forming a tree.
    \item \textbf{Note-taking} and \textbf{task-management tools} such as Org-mode for Emacs~\cite{OrgMode} or OmniOutliner~\cite{OmniOutliner} present the user with a tree structure that they can inspect and manipulate.
\end{itemize}

@@

In this paper we use an abstract model of a tree that can support any of the applications listed above, including distributed filesystems.
We allow replicas to manipulate this tree in any way: by creating new nodes, deleting nodes, or moving subtrees to a new location within the tree.
While there are many existing systems that support creating and deleting nodes (see Section~\ref{sec:relwork}), the key innovation of our algorithm is the support for moving subtrees.
We explain in Section~\ref{sec:move-is-hard} why this move operation is so challenging.
In summary, our contributions are:
\begin{itemize}
    \item We define a Conflict-free Replicated Data Type for trees that supports a \emph{move} operation, without requiring any coordination between replicas such as locking or consensus.
        As discussed in Section~\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
    \item We formalise an abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of the algorithm.
        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) while keeping the data in a tree structure.
    \item To demonstrate the practical viability of our approach, we refine our abstract algorithm to an efficient implementation within Isabelle/HOL and prove the equivalence of the two.  We thereafter extract a formally verified, executable Scala implementation from the Isabelle/HOL definitions, and perform tests on this implementation.
    \item We perform experiments with replicated filesystem products such as Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
\end{itemize}

@@

The first difficulty arises when a node is concurrently moved into different locations on different replicas.
This scenario is illustrated in Figure~\ref{fig:move-same-item}, where replica 1 moves node $A$ to be a child of $B$, while concurrently replica 2 moves $A$ to be a child of $C$.
After the replicas communicate these operations to each other, what should the merged state of the tree be?

@@

In an unreplicated tree it is easy to prevent cycles being introduced: if the node being moved is an ancestor of the destination node, then the operation is rejected.
However, a new problem arises in a replicated setting: different replicas may perform operations that are individually safe, but whose combination leads to a cycle.

@@

In this section we walk through the code in Figure~\ref{fig:code} step by step, explaining features of the Isabelle/HOL syntax as we encounter them.
Additional background documentation is available~\cite{DBLP:books/sp/NipkowK14}.

@@

We define only one kind of operation: \isa{Move t p m c} (Figure~\ref{fig:code}, lines 1--5).
A move operation is a record consisting of four fields: a timestamp \isa{t} of type \isa{'t}, a parent node \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node \isa{c} of type \isa{'n}.
Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be \emph{substituted} for arbitrary types; our only requirement is that timestamps \isa{'t} are globally unique and have an associated total order (e.g. Lamport timestamps~\cite{Lamport:1978jq} would suffice).

The meaning of an operation \isa{Move t p m c} is that at time \isa{t}, the child node \isa{c} is moved to be a child of parent node \isa{p}.
The operation does not specify the old location of \isa{c}; the algorithm simply ``steals'' \isa{c} from wherever it is currently located in the tree, and moves it to \isa{p}.
If \isa{c} does not currently exist in the tree, it is created as a child of \isa{p}.
For this reason we do not need a separate operation type to create tree nodes: simply performing a move with a fresh child node \isa{c} is sufficient to create \isa{c}.
Nor do we define a remove operation: instead, we can designate a particular tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.

@@

When a replica applies a \isa{Move} operation to its tree, it also records a corresponding \isa{LogMove} operation in its log.
The \isa{t}, \isa{p}, \isa{m} and \isa{c} fields are taken directly from the \isa{Move} record, while the \isa{oldp} field is filled in based on the state of the tree before the move.
If \isa{c} did not exist in the tree prior to the move, \isa{oldp} is set to \isa{None}.
Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}, that is: if there exists a \isa{p'} and \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.
The \isa{get\_parent} function implements this (lines 16--20).

@@

If the log is nonempty (line 48), we take \isa{logop} to be the first element of the log, and \isa{ops} to be the rest.
(The hash character in \isa{logop {\isacharhash} ops} is the \emph{cons} operator that adds one element to the head of a list.)
If the timestamp of \isa{logop} is greater than the timestamp of the new operation (line 49) we first undo \isa{logop} with \isa{undo\_op}, then recursively apply the new operation to the remaining log, and finally reapply \isa{logop} with \isa{redo\_op} (line 50).
Otherwise we perform the operation using \isa{do\_op}, and add the corresponding \isa{LogMove} record as the head of the log (line 51).
This logic ensures that the log is maintained in descending timestamp order, with the greatest timestamp at the head.
\isa{undo\_op} (lines 33--36) inverts the effect of a previous move operation by restoring the prior parent and metadata that were recorded in the \isa{LogMove}'s additional field.
\isa{redo\_op} (lines 38--41) uses \isa{do\_op} to perform an operation again and recomputes the \isa{LogMove} record (which might have changed due to the effect of the new operation with a lower timestamp).

@@

The first invariant we prove is that each tree node has either no parent (if it is the root of a tree) or exactly one parent (if it is a non-root node).
We state this theorem in Isabelle as follows, where \isa{apply\_ops\_unique\_parent} is the name we give to this theorem:
\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ apply{\isacharunderscore}ops{\isacharunderscore}unique{\isacharunderscore}parent{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isacharequal}\ {\isacharparenleft}log{\isacharcomma}\ tree{\isacharparenright}{\isacartoucheclose}\isanewline
\ \ \isakeyword{shows}\ {\isacartoucheopen}unique{\isacharunderscore}parent\ tree{\isacartoucheclose}
\end{isabelle}
\noindent That is, we consider any list of operations \isa{ops} and define \isa{(log, tree)} to be the replica state after \isa{ops} have been applied.
We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 56--57 of Figure~\ref{fig:code}: it states that whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.
As we make no assumptions about \isa{ops}, this theorem holds for any replica state that can be reached by applying any number of operations.
Proving \isa{apply\_ops\_unique\_parent} is easy because each of the \isa{do\_op}, \isa{undo\_op} and \isa{redo\_op} functions individually maintains the \isa{unique\_parent tree} invariant, and thus \isa{apply\_op} also maintains the invariant.

@@

To prove that \isa{apply\_op} preserves acyclicity, we need to show that for any tree produced by \isa{apply\_ops}, there exists a sequence of operations such that applying those operations using \isa{do\_op} invocations alone (not using \isa{undo\_op} or \isa{redo\_op}) results in the same tree.
(This is the case when operations are applied in order of increasing timestamp.)
Then we can use the result of the previous paragraph to prove that any tree produced by \isa{apply\_ops} must be acyclic.
